---
title: "Differential Expression"
author: "Insert author name"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 9
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
    code_folding: hide
    theme: "flatly"
    highlight: pygments
  includes:
    before_body: header.html
    after_body: footer.html
  css: style.css
---

````{r opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
````

<details><summary>Setup</summary>

* Libraries

````{r }
suppressPackageStartupMessages({
    library(data.table)
    library(DESeq2)
    library(emoji)
    library(gplots)
    library(here)
    library(hyperSpec)
    library(RColorBrewer)
    library(tidyverse)
    library(VennDiagram)
})
````

* Helper files

````{r }
suppressMessages({
    source(here("UPSCb-common/src/R/plotEnrichedTreemap.R"))
    source(here("UPSCb-common/src/R/featureSelection.R"))
    source(here("UPSCb-common/src/R/volcanoPlot.R"))
    source(here("UPSCb-common/src/R/functions_DifferentialExpression.R"))
})
````

* Graphics

````{r }
pal=brewer.pal(8,"Dark2")
hpal <- colorRampPalette(c("blue","white","red"))(100)
mar <- par("mar")
````


# DESeq2

All output files can be found in the DE_results folder, where files ending in "genes.csv" contain gene list which have been filtered based on a FDR threshold of 0.01 and a minumum log2FC of 0.5.

The files ending in "results.csv" contain all DE genes (without any filtering based on quality criteria). 

* Running the pipeline

````{r DEseq2, include=FALSE}
dds <- readRDS(file=here("data/analysis/salmon/dds_merged-tech-reps.rds"))

## IMPORTANT NOTE: the samples dataframe must have a column called "SampleID" and the values in SampleID must be the same of colnames(dds)

samples <- read_csv(here("doc/samples_merged-tech-reps.csv"),show_col_types = FALSE)

#Set the desired degisn
design(dds) <- formula(~ VARIABLE1 * VARIABLE2)

vsd <- varianceStabilizingTransformation(dds,blind=FALSE)
vst <- assay(vsd)
vst <- vst - min(vst)

dir.create(here("data/analysis/DE"),showWarnings=FALSE)
save(vst,file=here("data/analysis/DE/vst-aware.rda"))
write_delim(as.data.frame(vst) %>% rownames_to_column("ID"),
            here("data/analysis/DE/vst-aware.tsv"))

dds <- DESeq(dds)
resultsNames(dds)

````

<details><summary>Dispersion</summary>

The dispersion estimation is adequate.

````{r disp_est, fig.cap="Dispersion estimation"}
plotDispEsts(dds)
````

</details>

<details><summary>Results graphs per contrast</summary>

## Extracting results

Remember to remove suspicious DEGs from the results if you do not want to keep them

````{r contrasts}



#Reload dds object
dds <- readRDS(file=here("data/analysis/salmon/dds_merged-tech-reps.rds"))


# Use map to apply the function to each comparison
# INTEREST_COL is the name of a column with all the relevant information on a sample.
# variables_interest is a vector of column names with the names of all the columns included in the design.
# For example, if the design is "time * treatment", then the INTEREST_COL will be the name of a column containing values like:  time1_treatment1, time2_treatment1 and so on
# variables_interest will be instead a vector like this: c("time", "treatment")

# Example samplesheet
# Samp  | Time   | Treatment
# samp1 | spring | control
# samp2 | spring | drought
# samp3 | winter | control
# samp4 | winter | drought

# If the design is ~ Treatment, then Interest_Col is Treatment, and variables_interest is c("Treatment").
# Then the comparison is
# comparisons <- list(
#   c("drought",	"control"))
# It's recommended that the second element is the baseline, in this case = control

# If the design is ~ Time*Treatment, then you need to make a new column combining both together like
# Samp  | Time   | Treatment | Time_Treatment
# samp1 | spring | control   | spring_control
# samp2 | spring | drought   | spring_drought
# samp3 | winter | control   | winter_control
# samp4 | winter | drought   | winter_drought
# Then, Interest_Col is Time_Treatment, and variables_interest is c("Time","Treatment")
# Then the comparison is
# comparisons <- list(
#   c("spring_drought",	"spring_control"),
#   c("winter_drought",	"winter_control"),
#   c("spring_drought",	"winter_drought"))
# It's recommended that the second element is the baseline, in this case = spring_control, winter_control and winter_drought, respectively for our three comparisons

#Set the desired degisn
design(dds) <- formula(~ VARIABLE1 * VARIABLE2)

# Define your comparisons as a list of vectors, each with two values of INTEREST_COL that you wish to compare
comparisons <- list(
  c("Condition1",	"Condition2"),
  c("Condition1",	"Condition3"),
  c("Condition2",	"Condition3")
)


results_list <- map(comparisons, ~ process_comparison(.x, "INTEREST_COL", c("VARIABLE1", "VARIABLE2"), dds=dds, samples=samples, lfc=0.5, padj=0.01, default_dir=here("data/analysis/DE")))


# Set names for the results list
names(results_list) <- map_chr(comparisons, function(comparison) {
  gsub(" ", "", paste(comparison, collapse="vs"))
})
  

````

## Group DEGs in a single table

````{r grouping}

path <- here("data/analysis/DE/")
  
# Get a list of all csv files in the directory
files <- list.files(path, pattern = "*genes.csv")

# Initialize an empty data frame
all_data <- data.frame()

# Loop through the files
for(file in files){
  # Read the csv file
  data <- read_csv(paste0(path, "/", file), show_col_types = FALSE)
  
  
  # Add a new column with the file name
  data$Comparison <- file
  
  # Bind the data to the existing data frame
  all_data <- bind_rows(all_data, data)
}

# Split the 'Comparison' column using 'vs' as the separator
split_data <- strsplit(all_data$Comparison, split = "vs")

# Extract 'interest' and 'control' values
all_data$interest <- sapply(split_data, function(x) sub("DE-", "", x[1]))
all_data$control <- sapply(split_data, function(x) sub("_genes.csv", "", x[2]))

# Rename the first column
colnames(all_data)[colnames(all_data) == "...1"] <- "Gene"

write_tsv(all_data, here("data/analysis/DE/All_DEGs.tsv"))

# Save DE results list

saveRDS(results_list, file=here("data/analysis/DE/DE_results_list.rds"))

````

</details>

## DEGs number

This graph shows the number of differentially expressed genes in every situation.

````{r visualization}

library(reshape2)


# Convert the list of lists to a data frame
df <- do.call(rbind, lapply(seq_along(results_list), function(i) {
  data.frame(list = factor(names(results_list)[i], levels = names(results_list)),
             up = length(results_list[[i]]$up),
             dn = length(results_list[[i]]$dn))
}))

df_melt <- melt(df, id.vars = "list")

# Create the stacked bar plot

ggplot(df_melt, aes(x = list, y = value, fill = variable)) +
  geom_bar(stat = "identity") +
  labs(x = "Contrast", y = "Differentially expressed genes", fill="Differential expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

detach("package:reshape2", unload=TRUE)


````


## Compare selected conditions with venn diagrams and upset plots

### {.tabset .tabset-pills .unlisted .unnumbered}

#### Venn diagram 1 {-}

````{r }
grid.newpage()

list_for_Venn <- list(results_list[["Group1vsControl"]]$up,
                      results_list[["Group1vsControl"]]$dn,
                      results_list[["Group2vsControl"]]$up,
                      results_list[["Group2vsControl"]]$dn)
                      
                      
grid.draw(venn.diagram(list_for_Venn,
                       NULL,
                       fill=pal[1:4],
                       category=c("Group1vsControl_up",
                                  "Group1vsControl_dn",
                                  "Group2vsControl_up",
                                  "Group2vsControl_dn"
)))
````

#### Upset plot 1 {-}

Upset plots are alternatives to Venn diagrams, able to efficiently compare a higher number of conditions.
Use the nintersects option to set how many intersections you want to visualize (by default it is all of them). The intersections with the highest number of elements will be presented first, unless you change the order.by option

````{r }
library(UpSetR)

# Combine all DEG genes into a unique set
all_genes <- unique(all_data$Gene)


gene_matrix <- data.frame(
  gene = all_genes,
  C_T89vsSD_T89 = all_genes %in% results_list[["Group1vsControl"]]$up,
  LD_T89vsC_T89 = all_genes %in% results_list[["Group1vsControl"]]$dn,
  C_FPF1KOvsSD_FPF1KO = all_genes %in% results_list[["Group2vsControl"]]$up,
  LD_FPF1KOvsC_FPF1KO = all_genes %in% results_list[["Group2vsControl"]]$dn,
  C_FPF1OEvsSD_FPF1OE = all_genes %in% results_list[["Group2vsGroup1"]]$up,
  LD_FPF1OEvsC_FPF1OE = all_genes %in% results_list[["Group2vsGroup1"]]$dn
)


num_df <- gene_matrix %>% 
  dplyr::select(-gene) %>% 
  mutate(across(everything(), as.integer))

# Identify all-zero columns
all_zero_cols <- num_df %>%
  select(where(~ all(. == 0))) %>%
  names()

# Drop them
num_df <- num_df %>%
  select(where(~ any(. != 0)))

# Print them
if (length(all_zero_cols) > 0) {
cat("Dropped comparisons with no DEGs:", paste(all_zero_cols, collapse = ", "), "\n")
}


upset(num_df, 
      sets = colnames(num_df),   # Use column names as sets
      keep.order = TRUE,
      order.by = "freq",             # Order by frequency of intersection
      matrix.color = "#125223",     # Color for matrix points
      sets.bar.color = "#72b083",
      sets.x.label = "Comparisons",
      mainbar.y.label = "DEGs",
      nintersects = NA,
      point.size = 2.3,                # Adjust point size
      line.size = 0.8               # Adjust line size
)


````

# GO enrichment
For each comparison, GO enrichment is run on all the differentially expressed genes, but also on just the upregulated and the  downregulated ones, using a parent-child Fisher test with an FDR threshold of 0.05.


* Establish annotation and background

````{r prep_annotation}

# Loading functions necessary for TopGO run
source(here("UPSCb-common/src/R/functions_GOenrichment_TopGO.R"))

mapping <- "/mnt/picea/storage/reference/Arabidopsis-thaliana/TAIR10/gopher/tair10_gene_to_go.tsv"
annotation <- prepAnnot(mapping)
files <- list.files(path = here("data/analysis/DE/"), pattern = "results\\.csv$", full.names=TRUE)

background <- map(files, get_background) %>% 
  unlist() %>% 
  base::unique()

````


* Calculate the enrichment
````{r enrichment}
library(topGO)
library(purrr)

# Making a list of the vectors I want to test for enrichment
# By default, the upregulated and downregulated genes from every contrast are tested.
res <- list()
for (name in names(results_list)) {
  res[[paste0(name, "_all")]] <- results_list[[name]]$all
  res[[paste0(name, "_up")]] <- results_list[[name]]$up
  res[[paste0(name, "_dn")]] <- results_list[[name]]$dn
}


#Removing empty lists
res <- Filter(function(x) length(x) > 1, res)

# Apply topGO_combined to each element in res
results <- map(res, ~ topGO_combined(.x, background, annotation))

# Remove tibbles with 0 rows from results
results <- discard(results, ~ nrow(.x) == 0)

# Export each tibble in results to a TSV file
walk2(results, names(results), ~ write_tsv(.x, file = here(paste0("data/analysis/DE/GO-enr_", .y, ".tsv"))))





````

## Group GO enrichments in a single table

````{r grouping_enr}

path <- here("data/analysis/DE/")
  
# Get a list of all csv files in the directory
files <- list.files(path, pattern = "GO-enr.*.tsv")

# Initialize an empty data frame
all_data <- data.frame()

# Loop through the files
for(file in files){
  # Read the csv file
  data <- read_tsv(paste0(path, "/", file), show_col_types = FALSE)
  
  
  # Add a new column with the file name
  data$dataset <- file
  
  suppressMessages(
    data <- data %>%
    dplyr::mutate(FDR = ifelse(FDR == "< 1e-30", 0, FDR))
  )
  
  data$FDR <- as.numeric(data$FDR)
  
  # Bind the data to the existing data frame
  all_data <- bind_rows(all_data, data)
}



write_tsv(all_data, here("data/analysis/DE/All_DEGs_Enrichment.tsv"))

# Save GO results list

saveRDS(results, file=here("data/analysis/DE/GO_results_list.rds"))

````

## GO enrichment wordclouds

Note that the wordclouds can sometimes not show some of the GO terms, if their names could not be fit to the page.

To see all the enriched terms it is better to check the original tables in the DE folder.


`r emoji("point_right")` **Most significantly enriched terms appear bigger in the wordclouds.**


<details><summary>Wordclouds</summary>

````{r wordclouds}
library(wordcloud)

for (name in names(results)) {
  
    print(paste0("Enriched GO terms in ", name))

  suppressMessages(
    data <- results[[name]] %>%
    dplyr::mutate(FDR = ifelse(FDR == "< 1e-30", "1e-30", FDR)) %>%
    dplyr::mutate(scores = -log10(as.numeric(FDR))) %>% 
                    dplyr::select(Term, scores)
  )


  wordcloud(words = data$Term, 
            freq = data$scores, 
            min.freq = 0,           
            max.words = 200, 
            random.order = FALSE, 
            rot.per = 0, 
            scale = c(2.4,.3),  
            colors = brewer.pal(8, "Dark2"))
  }
  


````
</details>


## GO enrichment dotplots

`r emoji("point_right")` **The following plots show the top 20 most significantly enriched GO terms for each of the lists of upregulated or downregulated genes.**



<details><summary>Dotplots</summary>

````{r dotplots}
ntop <- 20

for (name in names(results)) {
  
  suppressMessages(
    data <- results[[name]] %>%
    dplyr::mutate(FDR = parse_double(sub("<","",FDR))) %>%
    dplyr::mutate(scores = -log10(as.numeric(FDR))) %>% 
                    dplyr::select(Term, scores)
  )

  
  # Add this line to filter the top terms
  data <- data %>% dplyr::arrange(desc(scores)) %>% dplyr::slice(1:ntop)
  
  # Convert Term to a factor and specify the levels to be in the order of descending scores
  data$Term <- factor(data$Term, levels = data$Term[order(data$scores, decreasing = TRUE)])
  
p <- ggplot(data,
  aes(x = Term, y = scores, size = scores, fill = scores)) +
  expand_limits(y = 1) +
  geom_point(shape = 21) +
  scale_size(range = c(2.5,12.5), name="-log10(FDR)") +
  scale_fill_continuous(low = 'royalblue', high = 'red4', name="-log10(FDR)") +
  xlab('') +
  ylab('-log10(FDR)') +
  labs(
    title = paste0("Enriched GOs for result ", name),
    subtitle = paste('Top', ntop, 'terms ordered by adjustes pvalue'),
    caption = 'Cut-off lines drawn at equivalents of p=0.05, p=0.01, p=0.001') +
  geom_hline(yintercept = c(-log10(0.05), -log10(0.01), -log10(0.001)),
    linetype = c("dotted", "longdash", "solid"),
    colour = c("black", "black", "black"),
    size = c(0.5, 1.5, 3)) +
  theme_bw(base_size = 24) +
  theme(
    legend.position = 'right',
    legend.background = element_rect(),
    plot.title = element_text(angle = 0, size = 16, face = 'bold', vjust = 1),
    plot.subtitle = element_text(angle = 0, size = 14, face = 'bold', vjust = 1),
    plot.caption = element_text(angle = 0, size = 12, face = 'bold', vjust = 1),

    axis.text.x = element_text(angle = 0, size = 12, face = 'bold', hjust = 1.10),
    axis.text.y = element_text(angle = 0, size = 12, face = 'bold', vjust = 0.5),
    axis.title = element_text(size = 12, face = 'bold'),
    axis.title.x = element_text(size = 12, face = 'bold'),
    axis.title.y = element_text(size = 12, face = 'bold'),
    axis.line = element_line(colour = 'black'),

    #Legend
    legend.key = element_blank(), # removes the border
    legend.key.size = unit(1, "cm"), # Sets overall area/size of the legend
    legend.text = element_text(size = 14, face = "bold"), # Text size
    title = element_text(size = 14, face = "bold")) +
  coord_flip()

print(p)
}

# ggplot2::ggsave(here("analysis/GO/A_vs_B_Fisher.pdf"),
#                device = NULL,
#                height = 8.5,
#                width = 12)
````
</details>


**Below are plot from top 20 most significantly enriched GO terms in each category:**

* BP: Biological Process

* CC: Cellular Component

* MF: Molecular Function
 

<details><summary>Dotplots</summary>

````{r dotplots2}
ntop <- 20

iwalk(results, function(e,r){
  dat <- e %>%
    mutate(FDR = parse_double(sub("<","",FDR)),
           GeneRatio = Significant/Annotated,
           Count = as.numeric(Significant)) %>%
    group_by(`GO category`) %>%
    arrange(FDR) %>%
    slice(1:ntop) %>%
    ungroup() %>%
    mutate(Term = factor(Term, levels = Term[order(GeneRatio)]))
  p <- ggplot(dat, aes(x =Term, y = GeneRatio, color = FDR, size = Count)) + 
    geom_point() +
    scale_color_gradient(low = "red", high = "blue") +
    theme_bw() + 
    ylab("GeneRatio") + 
    xlab("") + 
    facet_grid(`GO category` ~ ., scales = "free", space = "free") +
    ggtitle(paste0(r, ": "," GO enrichment")) +
    coord_flip()
  plot(p)
})
````
</details>

# Session Info
<details><summary>Setup</summary>
 ````{r session info, echo=FALSE}
 sessionInfo()
 ````
</details>
